---
title: "Assessing the Effect of Class Size on Math Scores of the First Graders"
output:
   html_document:
    df_print: paged
    fig_caption: yes
    number_sections: yes
---
# Introduction

# Data Description
The original data of 1st grade math scores were explored after removing all the missing data, resulting in 6598 observations with full information of class size, teacher ID and school ID. In order to make causal inference on the effect of class size on math score, Stable Unit Treatment Value Assumption (SUTVA) has to hold. Therefore, teacher (or class) was used as the unit of analysis instead of individual student, to avoid the interaction among students within one class. As shown in Figure 1, the distribution of math scores was nearly symmetric, and the median and mean scores for each class were very close as shown in Figure 2. Mean and median are both good summary statistics in this case, but average score is more often used to evaluate a class or a teacher. Therefore, mean math scores of students taught by each teacher were used for further data analysis. 

```{r, echo=FALSE, include=FALSE}
library(foreign)
star = read.spss("C:/Users/lgrey/Desktop/STA 207/Project 2/PROJECT+STAR/STAR_Students.sav",
                 to.data.frame = TRUE)
dat = star[c('g1schid', 'g1tchid', 'g1classtype', 'g1tmathss')]
dat1 = na.omit(dat)
library(knitr)
sum_sta = round(append(summary(dat1$g1tmathss)[c(1,6,3,4)], sd(dat1$g1tmathss)), 2)
names(sum_sta) = c('Minimum', 'Maximum', 'Median', 'Mean', 'SD')
```

__Table 1 Summary statistics of 1st grade math scores__

```{r, echo=FALSE}
kable(t(sum_sta))
```

```{r, echo=FALSE}
plot(density(dat1$g1tmathss), main = '', xlab = 'Math scores of 1st grade students')
```

__Figure 1 Distribution of 1st grade math scores__

```{r, echo=FALSE}
dat2 = aggregate(g1tmathss ~ g1tchid * g1classtype * g1schid, data = dat1, mean)
dat3 = table(dat1$g1tchid)
dat4 = aggregate(g1tmathss ~ g1tchid * g1classtype * g1schid, data = dat1, median)
plot(dat2$g1tmathss, type = 'l', col = 4, xlab = 'Teacher (or class)', ylab = 'Math score')
lines(dat4$g1tmathss, col = 2)
legend('topleft', legend = c('Mean', 'Meidan'), col = c(4, 2), lty = c(1, 1))
```

__Figure 2 Means and medians of maths scores for each teacher__

339 means math score of each teacher were obtained, with 3 class sizes in 76 different schools. As shown in Figure 3, the number of math score observed for each teacher ranged from 11 to 29, with most teachers having 20 to 24 observations (45 %). The distribution of mean math score is shown in Figure 3, which is also symmetric. The data obtained was unbalanced as each school had different number of each class size, which might cause unequal variances in data analysis. This issue will be discussed in the following.

```{r, echo=FALSE}
hist(dat3, main = '', xlab = 'Number of students taught by one teacher') 
```

__Figure 3 Histogram of the number of math score observed for each teacher__

```{r, echo=FALSE}
plot(density(dat2$g1tmathss), main = '', xlab = 'Mean math scores of students taught by each teacher ')
```

__Figure 4 Distribution of mean math score of students taught by each teacher__

