---
title: "Assessing the Effect of Class Size on Math Scores of the First Graders"
output:
  pdf_document: default
  html_document:
    df_print: paged
    fig_caption: yes
    number_sections: yes
---
Team ID: 9

Zhikuan Quan

Daidai Zhang

Wenfeng Chang

Jinghui Li

Github repo:

# Introduction

# Data Description
The original data of 1st grade math scores were explored after removing all the missing data, resulting in 6598 observations with full information of class size, teacher ID and school ID. The summary statistics of math score are shown in Table 1. Stable Unit Treatment Value Assumption (SUTVA) has to hold in order to make causal inference on the effect of class size on math score. Therefore, teacher (or class) was used as the unit of analysis instead of individual student, to avoid the interaction among students within one class. As shown in Figure 1, the median and mean scores for each class are very close. Mean and median are both good summary statistics in this case, but average score is more often used to evaluate a class or a teacher. Therefore, mean math scores of students taught by each teacher were calculated for further data analysis. 

```{r, echo=FALSE, include=FALSE}
library(foreign)
star = read.spss("C:/Users/lgrey/Desktop/STA 207/Project 2/PROJECT+STAR/STAR_Students.sav",
                 to.data.frame = TRUE)
dat = star[c('g1schid', 'g1tchid', 'g1classtype', 'g1tmathss')]
dat1 = na.omit(dat)
library(knitr)
sum_sta = round(append(summary(dat1$g1tmathss)[c(1,6,3,4)], sd(dat1$g1tmathss)), 2)
names(sum_sta) = c('Minimum', 'Maximum', 'Median', 'Mean', 'SD')
```

__Table 1 Summary statistics of 1st grade math scores__

```{r, echo=FALSE}
kable(t(sum_sta))
```


```{r, echo=FALSE}
dat2 = aggregate(g1tmathss ~ g1tchid * g1classtype * g1schid, data = dat1, mean)
dat3 = table(dat1$g1tchid)
dat4 = aggregate(g1tmathss ~ g1tchid * g1classtype * g1schid, data = dat1, median)
plot(dat2$g1tmathss, type = 'l', col = 4, xlab = 'Teacher (or class)', ylab = 'Math score')
lines(dat4$g1tmathss, col = 2)
legend('topleft', legend = c('Mean', 'Meidan'), col = c(4, 2), lty = c(1, 1))
```

__Figure 1 Means and medians of math scores for each teacher__

339 mean math scores of each teacher were obtained, with 3 class sizes in 76 different schools. The distribution of mean math score is shown in Figure 3, which is nearly symmetric. The data obtained was unbalanced as each school had different number of each class size, which might cause unequal variances in data analysis. This issue will be discussed in the following.


```{r, echo=FALSE}
plot(density(dat2$g1tmathss), main = '', xlab = 'Mean math scores of students taught by each teacher ')
```

__Figure 2 Distribution of mean math score of students taught by each teacher__

# Model Description

# Disccusion and Conclusion
This study and last study both found a significant effect of class size on 1st grade math score scores. However, unlike previous study, no difference was found between regular size class with aide and without aide, which could be due to the isolation of school effect in the current statistic model. 
